%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% UNIQUE
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\chapter{Unique Factorization of Ideals}

Unique factorization into irreducible elements frequently fails for
rings of integers of number fields.  In this chapter we will deduce a
central property of the ring of integers $\sO_K$ of an algebraic number
field, namely that every nonzero {\em ideal} factors uniquely as a
products of prime ideals.  Along the way, we will introduce fractional
ideals and prove that they form a free abelian group under
multiplication.  Factorization of {\em elements} of $\sO_K$ (and much
more!) is governed by the class group of $\sO_K$, which is the quotient
of the group of fractional ideals by the principal fractional ideals
(see Chapter~\ref{ch:classgroup}).

\begin{exercise}
  This exercise illustrates the failure of unique factorization in the ring
  $\sO_K$ of integers of $K = \QQ(\sqrt{-5})$.
  \begin{enumerate}
    \item Give an element $\alpha \in \sO_K$ that factors in two distinct
    ways into irreducible elements.
    \item Observe explicitly that the $(\alpha)$ factors uniquely, i.e.,
    the two distinct factorization in the previous part of this problem
    do not lead to two distinct factorization of the ideal $(\alpha)$
    into prime ideals.
  \end{enumerate}
  \begin{hint}
    $\sO_K = \ZZ[\sqrt{-5}]$.
  \end{hint}
\end{exercise}

\section{Dedekind Domains}

Recall (Corollary~\ref{prop:intnoetherian}) that we proved that the
ring of integers $\sO_K$ of a number field is noetherian as follows.
As we saw before using norms, the ring $\sO_K$ is finitely generated as
a module over~$\ZZ$, so it is certainly finitely generated as a ring
over~$\ZZ$.  By the Hilbert Basis Theorem (Theorem~\ref{thm:hilbert}),
$\sO_K$ is noetherian.

If $R$ is an integral domain, the \defn{field of fractions} $\Frac(R)$
of~$R$ is the field of all equivalence classes of formal quotients
$a/b$, where $a,b \in R$ with $b\neq 0$, and $a/b\sim c/d$ if $ad=bc$.
For example, the field of fractions of~$\ZZ$ is (canonically isomorphic
to) $\QQ$ and the field of fractions of $\ZZ[(1+\sqrt{5})/2]$ is
$\QQ(\sqrt{5})$.  The field of fractions of the ring $\sO_K$ of integers
of a number field~$K$ is just the number field~$K$ (see Lemma~\ref{lem:intq}).

\begin{example}
We compute the fraction fields mentioned above.
\begin{sagecode}
\begin{sagecell}
Frac(ZZ)
\end{sagecell}
\begin{sageout}
Rational Field
\end{sageout}
\end{sagecode}
In {\Sage} the {\tt Frac} command usually returns a field canonically
isomorphic to the fraction field (not a formal construction).
\begin{sagecode}
\begin{sagecell}
K.<a> = QuadraticField(5)
OK = K.ring_of_integers(); OK
\end{sagecell}
\begin{sageout}
Maximal Order in Number Field in a with defining polynomial x^2 - 5
\end{sageout}
\begin{sagecell}
OK.basis()
\end{sagecell}
\begin{sageout}
[1/2*a + 1/2, a]
\end{sageout}
\begin{sagecell}
Frac(OK)
\end{sagecell}
\begin{sageout}
Number Field in a with defining polynomial x^2 - 5
\end{sageout}
\end{sagecode} %link
\noindent{}The fraction field of an {\em order} -- i.e., a subring of $\sO_K$ of
finite index -- is also the number field again.
%link
\begin{sagecode}
\begin{sagecell}
O2 = K.order(2*a); O2
\end{sagecell}
\begin{sageout}
Order in Number Field in a with defining polynomial x^2 - 5
\end{sageout}
\begin{sagecell}
Frac(O2)
\end{sagecell}
\begin{sageout}
Number Field in a with defining polynomial x^2 - 5
\end{sageout}
\end{sagecode}
\end{example}

\begin{remark}
  Note that in computers {\tt 1/2 * x} means the same as {\tt (1/2)*x}.
  For more information about the order of operations in programming see
  \url{http://en.wikipedia.org/wiki/Order_of_operations}.
  In {\Sage} the {\tt \string^} symbol is replaced with python's
  exponentiation {\tt **} at execution.\footnote{
    Another source for order of operations specific to python is
    \url{https://docs.python.org/2/reference/expressions.html\#operator-precedence}.
  }
\end{remark}

\begin{definition}[Integrally Closed]
  An integral domain $R$ is \defn{integrally closed in its field of
  fractions} if whenever $\alpha$ is in the field of fractions of $R$
  and $\alpha$ satisfies a monic polynomial $f\in R[x]$, then $\alpha
  \in R$.
\end{definition}

For example, every field is integrally closed in its field of
fractions, as is the ring $\ZZ$ of integers.  However, $\ZZ[\sqrt{5}]$
is not integrally closed in its field of fractions, since
$(1+\sqrt{5})/2$ is integrally over $\ZZ$ and lies in $\QQ(\sqrt{5})$,
but not in $\ZZ[\sqrt{5}]$

\begin{proposition}\label{prop:integrallyclosed}\iprop{$\sO_K$ is integrally closed}
  If $K$ is any number field, then $\sO_K$ is integrally closed.  Also,
  the ring $\overline{\ZZ}$ of all algebraic integers (in a fixed choice of $\overline{\QQ}$)
  is integrally closed.
\end{proposition}
\begin{proof}
  We first prove that~$\overline{\ZZ}$ is integrally closed.  Suppose $\alpha\in\overline{\QQ}$
  is integral over~$\overline{\ZZ}$, so there is a monic polynomial $f(x)=x^n +
  a_{n-1}x^{n-1} + \cdots + a_1 x + a_0$ with $a_i\in\overline{\ZZ}$ and
  $f(\alpha)=0$.  The $a_i$ all lie in the ring of integers $\sO_K$ of the
  number field $K=\QQ(a_0,a_1,\dots a_{n-1})$, and $\sO_K$ is finitely
  generated as a $\ZZ$-module, so $\ZZ[a_0,\dots, a_{n-1}]$ is finitely
  generated as a $\ZZ$-module.  Since $f(\alpha)=0$, we can write $\alpha^n$
  as a $\ZZ[a_0,\dots, a_{n-1}]$-linear combination of $\alpha^i$ for $i<n$,
  so the ring $\ZZ[a_0,\dots, a_{n-1},\alpha]$ is also finitely generated
  as a $\ZZ$-module.  Thus $\ZZ[\alpha]$ is finitely generated as a $\ZZ$-module
  because it is a submodule of a finitely generated $\ZZ$-module, which
  implies that~$\alpha$ is integral over~$\ZZ$.

  Without loss we may assume that $K\subset \overline{\QQ}$, so that
  $\sO_K = \overline{\ZZ} \cap K$.
  Suppose $\alpha\in K$ is integral over $\sO_K$.  Then since $\overline{\ZZ}$ is
  integrally closed,~$\alpha$ is an element of $\overline{\ZZ}$, so
  $\alpha\in K \cap \overline{\ZZ}=\sO_K$, as required.
\end{proof}

\begin{exercise}\label{ex:Zbarnotnoetherian}
  Prove that $\overline{\ZZ}$ is not noetherian.

  \begin{hint}
    Find an ascending chain of ideals generated by prime fractional powers that does not stabilize.
  \end{hint}
\end{exercise}
%\begin{solution}
% Consider the chain of ideals $(2^{1/2}) \subsetneq (2^{1/4}) \subsetneq \cdots$.
% This is an ascending chain that does not stabilize, so $\overline{\ZZ}$ is not noetherian.
%\end{solution}

\begin{definition}[Dedekind Domain]
  An integral domain~$R$ is a \defn{Dedekind domain} if it is noetherian,
  integrally closed in its field of fractions, and every nonzero prime
  ideal of $R$ is maximal.
\end{definition}

\begin{exercise}
  Let $K$ be a field.
  \begin{enumerate}[label=(\alph*)]
    \item Prove that the polynomial ring $K[x]$ is a Dedekind domain.
    \item Is $\ZZ[x]$ a Dedekind domain?
  \end{enumerate}
\end{exercise}

The ring $\ZZ\oplus \ZZ$ is not a Dedekind domain because it is not an
integral domain.  The ring $\ZZ[\sqrt{5}]$ is not a Dedekind domain
because it is not integrally closed in its field of fractions.  The
ring~$\ZZ$ is a Dedekind domain, as is any ring of integers $\sO_K$ of a
number field, as we will see below.  Also, any field $K$ is a Dedekind
domain, since it is an integral domain, it is trivially integrally
closed in itself, and there are no nonzero prime ideals so the
condition that they be maximal is empty.

\begin{exercise}
  In Proposition~\ref{prop:integrallyclosed} we showed
  that $\overline{\ZZ}$ is integrally closed in its field of fractions.
  Prove that and every nonzero prime ideal of $\overline{\ZZ}$
  is maximal. Together with Exercise~\ref{ex:Zbarnotnoetherian},
  this shows $\overline{\ZZ}$ is not a Dedekind domain only because it
  is not noetherian.
\end{exercise}

\begin{exercise}\label{ex:dedekindlocal}
  Show that Dedekind domains are closed under
  localization. This means the following: given
  any nonzero prime $\p$ in $R$, the \emph{localization}
  $R_{\p}$ of $R$ at $\p$ is the ring formed by inverting all
  elements of $R$ not contained in $\p$. Thus $R_{\p}$ is a
  subring of the field of fractions $K$ of $R$ which contains
  $R$. For example, $\ZZ_{(2)}$ is the localization of $\ZZ$
  at the prime ideal $(2)$. Note $\ZZ_{(2)}$ contains
  $\frac{1}{3}$ but not $\frac{1}{2}$.
  This exercise will show $R_{\p}$ is again a Dedekind domain.
  In general, any element of $R_{\p}$ can be written as a quotient
  $\frac{a}{b}$ for some $a\in R$ and $b\in R\setminus \p$.

  \begin{hint}
    It is a standard fact of localizations that the
    set of prime ideals in $R_{\p}$ is in bijection with the set
    of prime ideals of $R$ contained in $\p$. Use this to show
    $R_{\p}$ is noetherian and all prime ideals of $R_{\p}$ are maximal.
    It remains to show $R_{\p}$ is integrally closed. Let $\alpha\in K$
    satisfy a monic polynomial with coefficients in $R_{\p}$. By
    clearing denominators show that $s\alpha\in R$ for some
    $s\in R\setminus \p$.
  \end{hint}
\end{exercise}

\begin{proposition}\iprop{$\sO_K$ is Dedekind}
  The ring of integers $\sO_K$ of a number field is a Dedekind domain.
\end{proposition}
\begin{proof}
  By Proposition~\ref{prop:integrallyclosed}, the ring $\sO_K$ is
  integrally closed, and by Proposition~\ref{prop:intnoetherian} it is
  noetherian.  Suppose that~$\p$ is a nonzero prime ideal of $\sO_K$.
  Let $\alpha\in \p$ be a nonzero element, and let $f(x)\in\ZZ[x]$ be the
  minimal polynomial of~$\alpha$.  Then
  $$f(\alpha)=\alpha^n+a_{n-1}\alpha^{n-1}+\cdots+a_1\alpha+a_0=0,$$ so
  $a_0 = -(\alpha^n+a_{n-1}\alpha^{n-1}+\cdots+a_1\alpha)\in \p$.  Since
  $f$ is irreducible, $a_0$ is a nonzero element of $\ZZ$ that lies
  in~$\p$.  Every element of the finitely generated abelian group
  $\sO_K/\p$ is killed by~$a_0$, so $\sO_K/\p$ is a finite set.
  Since~$\p$ is prime, $\sO_K/\p$ is an integral domain.  Every finite
  integral domain is a field (see Exercise~\ref{ex:finitedomain}), so
  $\p$ is maximal, which completes the proof.
\end{proof}

\begin{exercise}\label{ex:finitedomain}
  Prove that every finite integral domain is a field.
\end{exercise}

\section{Factorization of Ideals}

If $I$ and $J$ are ideals in a ring $R$, the product $IJ$ is the ideal
{\em generated by} all products of elements in $I$ with elements in $J$:
$$
  IJ = (ab : a\in I, b\in J) \subset R.
$$
Note that the set of all products $ab$, with $a\in I$ and $b\in J$,
need not be an ideal, so it is important to take the ideal generated
by that set.

\begin{exercise}\label{ex:idealprod}
  Give an example of two ideals $I, J$ in a commutative
  ring $R$ whose product is {\em not} equal to the set
  $\{ab : a \in I, b \in J\}$.
\end{exercise}

\begin{exercise}\label{ex:idealprod2}
  Suppose $R$ is a principal ideal domain. Is it always the case that
  $$
    IJ = \{ab : a \in I, b \in J\}
  $$
  for all ideals $I, J$ in $R$?
\end{exercise}


\begin{definition}[Fractional Ideal]\label{def:fracideal}
  A \defn{fractional ideal} is a nonzero $\sO_K$-submodule~$I$ of~$K$ that
  is finitely generated as an $\sO_K$-module.
\end{definition}

\begin{exercise}
  Is the set $\ZZ\left[\frac{1}{2}\right]$ of rational numbers with
  denominator a power of $2$ a fractional ideal?
\end{exercise}

We will sometimes call a genuine ideal $I\subset \sO_K$ an
\defn{integral ideal}.  The notion of fractional ideal makes
sense for an arbitrary Dedekind domain $R$ -- it is an
$R$-module $I\subset K=\Frac(R)$ that is finitely
generated as an $R$-module.

\begin{example}
  We multiply two fractional ideals in {\Sage}:
\begin{sagecode}
\begin{sagecell}
K.<a> = NumberField(x^2 + 23)
I = K.fractional_ideal(2, 1/2*a - 1/2)
J = I^2
I
\end{sagecell}
\begin{sageout}
Fractional ideal (2, 1/2*a - 1/2)
\end{sageout}
\begin{sagecell}
J
\end{sagecell}
\begin{sageout}
Fractional ideal (4, 1/2*a + 3/2)
\end{sageout}
\begin{sagecell}
I*J
\end{sagecell}
\begin{sageout}
Fractional ideal (1/2*a + 3/2)
\end{sageout}
\end{sagecode}
\end{example}

Since fractional ideals $I$ are finitely generated, we can clear
denominators of a generating set to see that there exists some nonzero
$\alpha\in K$ such that
$$
  \alpha I = J \subset \sO_K,
$$
with $J$ an integral ideal.  Thus dividing by $\alpha$, we see
that every fractional ideal is
of the form
$$
  a J = \{a b : b \in J\}
$$
for some $a\in K$ and integral ideal $J\subset \sO_K$.

For example, the set $\frac{1}{2}\ZZ$ of rational numbers with
denominator $1$ or $2$ is a fractional ideal of $\ZZ$.

\begin{theorem}\label{thm:ddabgrp}
  \ithm{fractional ideals}
  The set of fractional ideals of a Dedekind domain~$R$ is an
  abelian group under ideal multiplication with identity element $R$.
\end{theorem}
Note that fractional ideals are nonzero by definition, so it is not
necessary to write ``nonzero fractional ideals'' in the statement of
the theorem. We will {\em only} prove Theorem~\ref{thm:ddabgrp} in the
case when $R=\sO_K$ is the ring of integers of a number field $K$. The
general case can be found in many algebraic number theory books such
as \cite[Ch.~3]{marcus1977number}.
Before proving Theorem~\ref{thm:ddabgrp} we prove a lemma.  For the
rest of this section $\sO_K$ is the ring of integers of a number
field~$K$.


\begin{definition}[Divides for Ideals]
  Suppose that $I,J$ are ideals of $\sO_K$.
  Then we say that~$I$ \defn{divides}~$J$ if $I\supset J$.
\end{definition}
To see that this notion of divides is sensible, suppose $K=\QQ$, so
$\sO_K=\ZZ$.  Then $I=(n)$ and $J=(m)$ for some integer $n$ and $m$, and
$I$ divides $J$ means that $(n)\supset (m)$, i.e., that there exists
an integer~$c$ such that $m=cn$, which exactly means that $n$ divides
$m$, as expected.

\begin{lemma}\label{lem:divprod}\ilem{$I$ divides product of primes}
  Suppose~$I$ is a nonzero ideal of $\sO_K$.  Then there exist prime ideals
  $\p_1,\dots, \p_n$ such that $\p_1 \cdot \p_2 \cdots \p_n \subset I$,
  i.e.,~$I$ divides a product of prime ideals.
\end{lemma}
\begin{proof}
  Let $S$ be the set of nonzero ideals of $\sO_K$ that do not
  satisfy the conclusion
  of the lemma.  The key idea is to use that $\sO_K$ is noetherian to show that
  $S$ is the empty set.   If~$S$ is
  nonempty, then since $\sO_K$ is noetherian, there is an ideal
  $I\in S$ that is maximal as an element of~$S$.  If~$I$ were prime, then~$I$
  would trivially contain a product of primes, so we may assume that~$I$
  is not prime.  Thus there exists $a,b\in \sO_K$ such that $ab\in
  I$ but $a\not\in I$ and $b\not\in I$.  Let $J_1 = I+(a)$ and
  $J_2=I+(b)$.  Then neither $J_1$ nor $J_2$ is in $S$, since~$I$ is
  maximal, so both $J_1$ and $J_2$ contain a product of prime ideals,
  say $\p_1 \cdots \p_r \subset J_1$ and $\q_1\cdots \q_s \subset J_2$.
  Then
  $$
    \p_1  \cdots \p_r \cdot \q_1\cdots \q_s \subset
    J_1 J_2 = I^2 + I(b) + (a)I + (ab) \subset I,
  $$
  so $I$ contains a product of primes.  This is a contradiction,
  since we assumed $I\in S$.   Thus~$S$ is empty, which completes
  the proof.
\end{proof}
We are now ready to prove the theorem.

\begin{proof}[Proof of Theorem~\ref{thm:ddabgrp}]
  Note that we will {\em only} prove Theorem~\ref{thm:ddabgrp} in the
  case when $R=\sO_K$ is the ring of integers of a number field $K$.

  The product of two fractional ideals is again finitely generated, so
  it is a fractional ideal, and $I\sO_K=I$ for any ideal~$I$,
  so to prove that the set of fractional ideals under multiplication is
  a group it suffices to show the existence of inverses.  We will first
  prove that if~$\p$ is a prime ideal, then~$\p$ has an inverse, then we
  will prove that all nonzero integral ideals have inverses, and finally
  observe that every fractional ideal has an inverse.  (Note: Once we
  know that the set of fractional ideals is a group, it will follows
  that inverses are unique; until then we will be careful to write
  ``an'' instead of ``the''.)

  Suppose $\p$ is a nonzero prime ideal of $\sO_K$.   We will show that
  the $\sO_K$-module
  $$
    I = \{a \in K : a\p \subset \sO_K \}
  $$
  is a fractional ideal of $\sO_K$ such that $I\p = \sO_K$, so that
  $I$ is an inverse of $\p$.

  For the rest of the proof, fix a nonzero element $b\in \p$.  Since~$I$
  is an $\sO_K$-module, $bI\subset \sO_K$ is an $\sO_K$ ideal, hence~$I$ is
  a fractional ideal.  Since $\sO_K \subset I$ we have $\p \subset I \p
  \subset \sO_K$, hence since~$\p$ is maximal, either $\p = I\p$ or $I\p
  = \sO_K$.  If $I\p=\sO_K$, we are done since then~$I$ is an inverse
  of~$\p$.  Thus suppose that $I\p=\p$.  Our strategy is to show that
  there is some $d\in I$, with $d\not\in\sO_K$.  Since $I\p=\p$, such
  a~$d$ would leave~$\p$ invariant, i.e., $d \p \subset \p$.  Since $\p$
  is a finitely generated $\sO_K$-module we will see that
  it will follow that $d\in \sO_K$, a contradiction.

  By Lemma~\ref{lem:divprod}, we can choose a product $\p_1,\dots, \p_m$,
  with~$m$ minimal, with
  $$
    \p_1\p_2\cdots \p_m \subset (b) \subset \p.
  $$ If no $\p_i$ is contained in $\p$, then we can choose for each $i$
  an $a_i \in \p_i$ with $a_i\not\in \p$; but then $\prod a_i\in \p$,
  which contradicts that $\p$ is a prime ideal.  Thus some $\p_i$, say
  $\p_1$, is contained in $\p$, which implies that $\p_1 = \p$ since
  every nonzero prime ideal is maximal.  Because~$m$ is minimal,
  $\p_2\cdots \p_m$ is not a subset of $(b)$, so there exists $c\in
  \p_2\cdots \p_m$ that does not lie in $(b)$. Then $\p(c) \subset (b)$,
  so by definition of $I$ we have $d=c/b\in I$.  However, $d\not\in
  \sO_K$, since if it were then $c$ would be in $(b)$.  We have thus
  found our element $d\in I$ that does not lie in $\sO_K$.

  To finish the proof that $\p$ has an inverse, we observe that $d$
  preserves the finitely generated $\sO_K$-module $\p$, and is hence in
  $\sO_K$, a contradiction.
  More precisely, if $b_1,\dots, b_n$ is a basis for
  $\p$ as a $\ZZ$-module, then the action of~$d$ on~$\p$ is given by a
  matrix with entries in~$\ZZ$, so the minimal polynomial of~$d$ has
  coefficients in~$\ZZ$ (because $d$ satisfies the minimal polynomial of
  $\ell_d$, by the Cayley-Hamilton theorem -- here we also use that
  $\QQ\otimes\p = K$, since $\sO_K/\p$ is a finite set).  This implies
  that~$d$ is integral over~$\ZZ$, so $d\in \sO_K$ since $\sO_K$ is by
  definition the set of elements of $K$ that are integral over $\ZZ$.

  So far we have proved that if $\p$ is a prime ideal of $\sO_K$, then
  $$
    \p^{-1} = \{a \in K : a\p \subset \sO_K\}
  $$
  is the inverse of $\p$ in
  the monoid of nonzero fractional ideals of $\sO_K$.  As mentioned after
  Definition~\ref{def:fracideal}, every nonzero fractional
  ideal is of the form $aI$ for $a\in K$ and $I$ an integral ideal, so
  since $(a)$ has inverse $(1/a)$, it suffices to show that every
  integral ideal~$I$ has an inverse.  If not, then there is a nonzero
  integral ideal~$I$ that is maximal among all nonzero integral ideals
  that do not have an inverse.  Every ideal is contained in a maximal
  ideal, so there is a nonzero prime ideal $\p$ such that $I\subset \p$.
  Multiplying both sides of this inclusion by $\p^{-1}$ and using that
  $\sO_K\subset \p^{-1}$, we see that
  $$
    I \subset \p^{-1} I \subset \p^{-1}\p = \sO_K.
  $$
  If $I = \p^{-1} I$, then arguing as in the proof that $\p^{-1}$ is an
  inverse of~$\p$, we see that each element of $\p^{-1}$ preserves the
  finitely generated $\ZZ$-module~$I$ and is hence integral.  But then
  $\p^{-1}\subset \sO_K$, which, upon multiplying both sides by $\p$,
  implies that $\sO_K = \p \p^{-1} \subset \p$, a contradiction.
  Thus $I \neq \p^{-1} I$.  Because $I$ is
  maximal among ideals that do not have an inverse, the ideal $\p^{-1}
  I$ does have an inverse~$J$. Then $\p^{-1}J$ is an inverse of~$I$,
  since $(J\p^{-1})I = J(\p^{-1}I) = \sO_K$.
\end{proof}

We can finally deduce the crucial Theorem~\ref{thm:intuniqfac}, which
will allow us to show that any nonzero ideal of a Dedekind domain can
be expressed uniquely as a product of primes (up to order).  Thus
unique factorization holds for ideals in a Dedekind domain, and it is
this unique factorization that initially motivated the introduction of
ideals to mathematics over a century ago.

\begin{theorem}\label{thm:intuniqfac}\ithm{unique ideal factorization}
  Suppose $I$ is a nonzero integral ideal of $\sO_K$.  Then $I$ can
  be written as a product
  $$
    I = \p_1\cdots \p_n
  $$
  of prime ideals of $\sO_K$, and this representation is unique up to
  order.
\end{theorem}
\begin{proof}
  Suppose~$I$ is an ideal that is maximal among the set of all ideals in
  $\sO_K$ that cannot be written as a product of primes.  Every ideal is
  contained in a maximal ideal, so $I$ is contained in a nonzero prime
  ideal $\p$.  If $I\p^{-1} = I$, then by Theorem~\ref{thm:ddabgrp} we
  can cancel $I$ from both sides of this equation to see that
  $\p^{-1}=\sO_K$, a contradiction.  Since $\sO_K\subset \p^{-1}$,
  we have $I\subset I\p^{-1}$, and by the above observation~$I$
  is strictly contained in $I\p^{-1}$.
  By our maximality assumption on~$I$, there are maximal
  ideals $\p_1,\dots,\p_n$ such that $I\p^{-1} = \p_1\cdots \p_n$.
  Then $I=\p\cdot \p_1\cdots \p_n$, a contradiction.  Thus every ideal
  can be written as a product of primes.

  Suppose $\p_1\cdots \p_n=\q_1\cdots \q_m$. If no $\q_i$ is contained in
  $\p_1$, then for each $i$ there is an $a_i\in \q_i$ such that
  $a_i\not\in\p_1$.  But the product of the $a_i$ is in $\p_1\cdots
  \p_n$, which is a subset of $\p_1$, which contradicts that
  $\p_1$ is a prime ideal.  Thus $\q_i=\p_1$ for some~$i$.  We can thus
  cancel $\q_i$ and $\p_1$ from both sides of the equation by multiplying
  both sides by the inverse.
  Repeating this argument finishes the proof of uniqueness.
\end{proof}

\begin{exercise}
  Factor the ideal $(10)$ as a product of primes
  in the ring of integers of $\QQ(\sqrt{11})$. You are allowed
  to use a computer, as long as you show the commands you use.
  \begin{hint}
    In {\Sage}, an ideal {\tt I} can be factored using {\tt I.factor()}.
  \end{hint}
\end{exercise}

\begin{theorem}\label{thm:uniqfac}
  \icor{factorization of fractional ideals}
  If $I$ is a fractional ideal of $\sO_K$ then there exists
  prime ideals $\p_1,\dots,\p_n$ and $\q_1,\dots,\q_m$,
  unique up to order, such that
  $$
    I = (\p_1\cdots \p_n)(\q_1\cdots \q_m)^{-1}.
  $$
\end{theorem}
\begin{proof}
  We have $I=(a/b)J$ for some $a,b\in\sO_K$ and integral ideal $J$.
  Applying Theorem~\ref{thm:intuniqfac} to $(a)$, $(b)$, and $J$ gives
  an expression as claimed.  For uniqueness, if one has two such product
  expressions, multiply through by the denominators and use the
  uniqueness part of Theorem~\ref{thm:intuniqfac}.
\end{proof}

\begin{example}
  The ring of integers of $K=\QQ(\sqrt{-6})$ is $\sO_K=\ZZ[\sqrt{-6}]$.
  We have
  $$
    6 = -\sqrt{-6}\sqrt{-6} = 2 \cdot 3.
  $$
  If $ab=\sqrt{-6}$, with $a,b\in \sO_K$ and neither a unit, then
  $\Norm(a)\Norm(b) = 6$, so without loss $\Norm(a)=2$ and
  $\Norm(b)=3$. If $a=c + d\sqrt{-6}$, then $\Norm(a) = c^2 + 6d^2$;
  since the equation $c^2 + 6d^2 = 2$ has no solution with $c,d\in\ZZ$,
  there is no element in $\sO_K$ with norm~$2$, so $\sqrt{-6}$ is
  irreducible.  Also, $\sqrt{-6}$ is not a unit times~$2$ or times $3$,
  since again the norms would not match up.  Thus $6$ cannot be written
  uniquely as a product of irreducibles in $\sO_K$.
  Theorem~\ref{thm:uniqfac}, however, implies that the principal
  ideal $(6)$ can, however, be
  written uniquely as a product of prime ideals.
  An explicit decomposition is
  \begin{equation}\label{eqn:fac6}
    (6) = (2, 2+\sqrt{-6})^2 \cdot (3,3+\sqrt{-6})^2,
  \end{equation}
  where each of the ideals $(2, 2+\sqrt{-6})$ and $(3, 3+\sqrt{-6})$ is
  prime.  We will discuss algorithms for computing such a decomposition
  in detail in Chapter~\ref{ch:factoring_primes}.  The first idea is to
  write $(6)=(2)(3)$, and hence reduce to the case of writing the $(p)$,
  for $p\in\ZZ$ prime, as a product of primes.  Next one decomposes the
  finite (as a set) ring $\sO_K/p\sO_K$.

  The factorization (\ref{eqn:fac6}) can be compute using {\Sage} as follows:
\begin{sagecode}
\begin{sagecell}
K.<a> = NumberField(x^2 + 6); K
\end{sagecell}
\begin{sageout}
Number Field in a with defining polynomial x^2 + 6
\end{sageout}
\begin{sagecell}
K.factor(6)
\end{sagecell}
\begin{sageout}
(Fractional ideal (2, a))^2 * \
(Fractional ideal (3, a))^2
\end{sageout}
\end{sagecode}
\end{example}

\begin{exercise}
  Let $\sO_K$ be the ring of integers of a number field.
  Let $F_K$ denote the abelian group of fractional ideals of $\sO_K$.
  \begin{enumerate}[label=(\alph*)]
    \item Prove that $F_K$ is torsion free.
    \item Prove that $F_K$ is not finitely generated.
    \item Prove that $F_K$ is countable.
    \item Conclude that if $K$ and $L$ are number fields, then there
    exists some (non-canonical) isomorphism of groups $F_K \ncisom F_L$.
  \end{enumerate}
\end{exercise}
%  \begin{solution}
%    \begin{enumerate}
%      \item If $I$ is a fractional ideal, then $\Norm(I^k) = \Norm(I)^k$. So if $\Norm(I) \neq 1$ then $\Norm(I^k) \neq 1$ for any nonzero $k \in \ZZ$.
%      \item Let $S$ be a finite subset of $F_K$.
%      Choose a prime $p$ not dividing $\Norm(I)$ for any $I \in S$.
%      Since $\Norm(IJ) = \Norm(I)\Norm(J)$, $p$ also does not divide any ideal generated by $S$.
%      Hence $S$ does not generate the ideal $p\sO_K$.
%      \item This follows from Theorem~\ref{thm:uniqfac} and the countability of primes.
%      \item Every countably infinite free abelian group is isomorphic to $\bigoplus \ZZ$.
%    \end{enumerate}
%  \end{solution}

\begin{exercise}
  Give an example of each of the following, with proof:
  \begin{enumerate}[label=(\alph*)]
    \item A non-principal ideal in a ring.
    \item A module that is not finitely generated.
    \item The ring of integers of a number field of degree~$3$.
    \item An order in the ring of integers of a number field of degree~$5$.
    \item The matrix of left multiplication by an element of~$K$,
    where~$K$ is a degree~$3$ number field.
    You will first need to choose a basis.
    \item An integral domain that is not integrally closed in its field
    of fractions.
    \item A Dedekind domain with finite cardinality.
    \item A fractional ideal of the ring of integers of a number
    field that is not an integral ideal.
  \end{enumerate}
\end{exercise}
%\begin{solution}
%\begin{enumerate}
%  \item $(2,x)$ in $\ZZ[x]$.
%  \item $\ZZ[x_1,x_2,\dots]$ as a $\ZZ$-module.
%  \item $\sO_{\QQ(\sqrt[3]{2})} = \ZZ[\sqrt[3]{2}]$.
%  \item $\ZZ[\sqrt[5]{2}]$.
%  \item Multiplication by $\sqrt[3]{2}$ on $\QQ(\sqrt[3]{2})$ with basis $\{1,\sqrt[3]{2},(\sqrt[3]{2})^2\}$ is \begin{pmatrix} 0 & 0 & 2 \\ 1 & 0 & 0 \\ 0 & 1 & 0 \end{pmatrix}.
%  \item $\ZZ[\sqrt{5}]$.
%  \item $\FF_2$
%  \item $\frac{1}{2}\ZZ$.
%\end{enumerate}
%\end{solution}
